---
title: "Untitled"
author: "Andrey Koval"
date: "Monday, January 12, 2015"
output: html_document
---

<!--  Set the working directory to the repository's base directory; this assumes the report is nested inside of only one directory.-->
```{r, echo=F, message=F} 
library(knitr)
opts_knit$set(root.dir='../../../')  #Don't combine this call with any other chunk -especially one that uses file paths.
```

<!-- Set the report-wide options, and point to the external script file. -->
```{r, echo=F, message=T}
require(knitr)
opts_chunk$set(
  results='show', 
  message = TRUE,
  comment = NA, 
  tidy = FALSE,
#   fig.height = 4.8, 
#   fig.width = 6.5, 
  out.width = NULL,
  fig.path = 'figure_rmd/',     
  dev = "png",
  dpi = 70
)
echoChunks <- FALSE
warningChunks<- FALSE
messageChunks<- FALSE
outwidthChunks <- "90%"
options(width=120) #So the output is 50% wider than the default.
read_chunk("./Chapters/02/R_Chapter2/hoffman-ch2.R") # the file to which knitr calls for the chunks
```


## Beginnings of modeling

There are several packages in R suited to fit linear and non-linear models. Some of them that we will use are:  
 - <code> stats </code>   
 - <code> nlme  </code>   
 - <code> lme4  </code>   
 
 
### <code>stats</code>  package  
 
  - [Official Documentation](http://www.rdocumentation.org/packages/stats)  
  - [lm()](http://www.rdocumentation.org/packages/stats/functions/lm) function    
  - [glm()](http://www.rdocumentation.org/packages/stats/functions/glm) function     
  
  
### <code>nlme</code>  package  
  - [Official Documentation](http://www.rdocumentation.org/packages/nlme)  
  - [gls()](http://www.rdocumentation.org/packages/nlme/functions/gls) function    
  - see a basic example of model result processing in the slides on [Statistical Modeling](http://ialsa.github.io/COAG-colloquium-2014F/2014-11-18-Statistical-Modeling.html#35) of the COAG [Colloquium series](http://ialsa.github.io/COAG-colloquium-2014F) on reproducible research.  
  

### <code>lme4</code>  package  
  - [Official Documentation](http://www.rdocumentation.org/packages/lme4)  
  - [lmer](http://www.rdocumentation.org/packages/nlme/functions/gls) function 
  - [lmer guide](http://htmlpreview.github.io/?https://github.com/andkov/Longitudinal_Models_of_Religiosity_NLSY97/blob/master/Vignettes/lmer/for%20Appendix/lmerGuide.html#adding-model-output)
  

  

  
 
