---
title: "Untitled"
author: "Andrey Koval"
date: "Monday, January 12, 2015"
output: html_document
---

<!--  Set the working directory to the repository's base directory; this assumes the report is nested inside of only one directory.-->
```{r, echo=F, message=F} 
library(knitr)
opts_knit$set(root.dir='../../../')  #Don't combine this call with any other chunk -especially one that uses file paths.
```

<!-- Set the report-wide options, and point to the external script file. -->
```{r, echo=F, message=T}
require(knitr)
opts_chunk$set(
  results='show', 
  message = TRUE,
  comment = NA, 
  tidy = FALSE,
#   fig.height = 4.8, 
#   fig.width = 6.5, 
  out.width = NULL,
  fig.path = 'figure_rmd/',     
  dev = "png",
  dpi = 70
)
echoChunks <- FALSE
warningChunks<- FALSE
messageChunks<- FALSE
outwidthChunks <- "90%"
options(width=120) #So the output is 50% wider than the default.
read_chunk("./Chapters/05/R_Chapter5/hoffman-ch5.R") # the file to which knitr calls for the chunks
```


## Resources

#### General R


- [A GOOD PLACE TO START LEARNING R](http://www.rstudio.com/resources/training/online-learning/) - The RStudio team collects the best online resources.  

- [Swirl](http://swirlstats.com/students.html) and [DataCamp](https://www.datacamp.com/courses/introduction-to-r)  offer interactive courses.  

- for brief reviews of key books and resources see Will Beasley's [Resources Opinions](https://github.com/OuhscBbmc/RedcapExamplesAndPatterns/blob/master/DocumentationGlobal/ResourcesOpinions.md)  

 - another presentation by Will provides an excellent overview of [Statistical Collaboration with GitHub](http://htmlpreview.github.io/?https://raw.githubusercontent.com/OuhscBbmc/StatisticalComputing/master/2014_Presentations/05_May/BeasleyScugGitHub2014-05.html#/)
 - Winston Chan's [R Cookbook](http://shop.oreilly.com/product/9780596809164.do) is a perfect book to get you started with producing graphs with RStudio  
 - [Quick-R](http://www.statmethods.net/) Quick and convenient resource for R reference  
 - [60+ resources for learning R](http://www.computerworld.com/article/2497464/business-intelligence-60-r-resources-to-improve-your-data-skills.html) Will give you more than you can handle. Browse and see which resource fits your learning style and organizes information in the manner that makes sense to you.

#### Modeling

  - [stats::lm()](http://www.rdocumentation.org/packages/stats/functions/lm) function, see [Gelman & Hill](http://www.stat.columbia.edu/~gelman/arm/), pages 38-39       
  - [stats::glm()](http://www.rdocumentation.org/packages/stats/functions/glm) function     
  - [nlme::gls()](http://www.rdocumentation.org/packages/nlme/functions/gls) function    
  - see a basic example of model result processing in the slides on [Statistical Modeling](http://ialsa.github.io/COAG-colloquium-2014F/2014-11-18-Statistical-Modeling.html#35) of the COAG [Colloquium series](http://ialsa.github.io/COAG-colloquium-2014F) on reproducible research.  
  - [lme4::lmer](http://www.rdocumentation.org/packages/nlme/functions/gls) function,  see [Gelman & Hill](http://www.stat.columbia.edu/~gelman/arm/), pages 259-262 and Appendix C. 
  - [lmer guide](http://htmlpreview.github.io/?https://github.com/andkov/Longitudinal_Models_of_Religiosity_NLSY97/blob/master/Vignettes/lmer/for%20Appendix/lmerGuide.html#adding-model-output). Shows direct interaction with the model object, not through functions like Gelman & Hill show.




  

  
 
